Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-10-17 src/tex/ms.tex
Update ms.tex
Finalized all changes

\usepackage{showyourwork}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage[nolist,nohyperlinks]{acronym}
\usepackage{bookmark}
\usepackage{changes}

@@

\begin{abstract}
    The propagation of gravitational waves can reveal fundamental features of the structure of spacetime.
    For instance, differences in the propagation of gravitational-wave polarizations would be a smoking gun for parity violations in the gravitational sector, as expected from birefringent theories like Chern-Simons gravity.
    Here we look for evidence of amplitude birefringence in the \replaced{third catalog of detections by the Laser Interferometer Gravitational Wave Observatory and Virgo}{latest Laser Interferometer Gravitational Wave Observatory-Virgo catalog (GWTC-3)} through the use of birefringent templates inspired by dynamical Chern-Simons gravity.
    From 71 binary-black-hole signals, we obtain the most precise constraints on gravitational-wave amplitude birefringence yet, measuring a birefringent attenuation of \variable{output/restricted_kappa_median.txt} at $100 \, \mathrm{Hz}$ with 90\% credibility, equivalent to a parity-violation energy scale of \variable{output/M_PV_constraint.txt}.
\end{abstract}

@@

% Section guide
In Sec.~\ref{sec:Background}, we briefly review the background of \ac{GW} amplitude birefringence.
In Sec.~\ref{sec:Method}, we describe the modification we apply to the baseline \ac{GR} waveform, summarize our \ac{PE} configuration and outline our statistical methods for combining information from multiple events.
In Sec.~\ref{sec:Results}, we present our constraint on \ac{GW} amplitude birefringence, discussing individual events and the catalog collectively.
In Sec.~\ref{sec:Discussion}, we discuss the implications of our result, comparing to previous constraints in the literature and outlining correlation structures that appear in our measurements.
Finally, we conclude in Sec.~\replaced{\ref{sec:Conclusion}}{\ref{sec:Discussion}} with a summary.
\added{In the Appendix, we provide extended results and discussion for two notable events (GW170818 and GW190521).}

@@

Equation \eqref{eq:waveform_modification} can be derived as the first-order effect in an expansion away from \ac{GR} under multiple frameworks.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudoscalar field and its derivative in Chern-Simons gravity \cite{Alexander:2009tp}.
Since it originates from a truncated series expansion,%
\footnote{Often the birefringent effect is written as an expansion in redshift $z$ \cite[e.g.][]{Zhao:2019xmm}, rather than distance. Here we use $d_C$ to emphasize that the effect accumulates \emph{per cycle}, or with propagation distance.  Thus, current ground-based experiments, which can detect gravitational waves to $z \simeq 1$, $d_C \simeq d_H$ (Hubble distance), are already observing propagation over a large fraction of the Universe and constrain this effect meaningfully.} %
Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, namely
\begin{equation}
    \left(\frac{\left|\kappa\right|}{\mathrm{Gpc}^{-1}}\right) \left(\frac{d_c}{\mathrm{Gpc}}\right) \left(\frac{f}{100\, \mathrm{Hz}}\right) < 1\,,
    \label{eq:small_exponent}
\end{equation}
recalling that $\kappa$ has dimensions of inverse length.
Otherwise, more frequency-dependent terms could enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        Illustration of amplitude birefringence. The GR waveform for the $\ell=|m|=2$ mode of a nonprecessing BBH seen edge-on ($\cos\iota = 0$) is linearly polarized and thus contains equal amounts of left- and right-handed modes for all frequencies (dotted, top).
        However, if spacetime were birefringent \added{(BR)} following Eq.~\protect\eqref{eq:waveform_modification}, then the waveform observed on Earth would contain different fractions of the two circular modes, with higher frequencies affected more strongly (solid, top). 
        In the time domain, this manifests as a time-dependent amplification of the waveform, with a stronger effect at later times when the chirp reaches a higher instantaneous frequency (bottom).
        For this example, the black holes do not spin and have equal masses $m_1 = m_2 = 10\, M_\odot$, and we have chosen a luminosity distance $d_L = 400\, {\rm Mpc}$ and $\kappa = 0.6$.
        }
    \label{fig:birefringence}
\end{figure}

@@

To this end, we perform three additional \ac{PE} runs for GW200129, considering only two detectors at a time: LIGO Hanford and Virgo (\replaced{HV}{HVL}), LIGO Livingston and Virgo (LV), and LIGO Hanford and LIGO Livingston (HL).
If the preference for $\kappa < 0$ is tied to the glitch in Virgo, we expect it to disappear in the HL run, which excludes Virgo data.

>>>>

\usepackage{showyourwork}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage[nolist,nohyperlinks]{acronym}
\usepackage{bookmark}

@@

\begin{abstract}
    The propagation of gravitational waves can reveal fundamental features of the structure of spacetime.
    For instance, differences in the propagation of gravitational-wave polarizations would be a smoking gun for parity violations in the gravitational sector, as expected from birefringent theories like Chern-Simons gravity.
    Here we look for evidence of amplitude birefringence in the third catalog of detections by the Laser Interferometer Gravitational Wave Observatory and Virgo through the use of birefringent templates inspired by dynamical Chern-Simons gravity.
    From 71 binary-black-hole signals, we obtain the most precise constraints on gravitational-wave amplitude birefringence yet, measuring a birefringent attenuation of \variable{output/restricted_kappa_median.txt} at $100 \, \mathrm{Hz}$ with 90\% credibility, equivalent to a parity-violation energy scale of \variable{output/M_PV_constraint.txt}.
\end{abstract}

@@

% Section guide
In Sec.~\ref{sec:Background}, we briefly review the background of \ac{GW} amplitude birefringence.
In Sec.~\ref{sec:Method}, we describe the modification we apply to the baseline \ac{GR} waveform, summarize our \ac{PE} configuration and outline our statistical methods for combining information from multiple events.
In Sec.~\ref{sec:Results}, we present our constraint on \ac{GW} amplitude birefringence, discussing individual events and the catalog collectively.
In Sec.~\ref{sec:Discussion}, we discuss the implications of our result, comparing to previous constraints in the literature and outlining correlation structures that appear in our measurements.
Finally, we conclude in Sec.~\ref{sec:Discussion} with a summary.
In the Appendix, we provide extended results and discussion for two notable events (GW170818 and GW190521).

@@

Equation \eqref{eq:waveform_modification} can be derived as the first-order effect in an expansion away from \ac{GR} under multiple frameworks.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudoscalar field and its derivative in Chern-Simons gravity \cite{Alexander:2009tp}.
Since it originates from a truncated series expansion,%
\footnote{Often the birefringent effect is written as an expansion in redshift $z$ \cite[e.g.][]{Zhao:2019xmm}, rather than distance. Here we use $d_C$ to emphasize that the effect accumulates \emph{per cycle}, or with propagation distance.  Thus, current ground-based experiments, which can detect gravitational waves to $z \simeq 1$, $d_C \simeq d_H$ (Hubble distance), are already observing propagation over a large fraction of the Universe and constrain this effect meaningfully.} %
Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, namely
\begin{equation}
    \left(\frac{\left|\kappa\right|}{\mathrm{Gpc}^{-1}}\right) \left(\frac{d_c}{\mathrm{Gpc}}\right) \left(\frac{f}{100\, \mathrm{Hz}}\right) < 1\,,
    \label{eq:small_exponent}
\end{equation}
recalling that $\kappa$ has dimensions of inverse length.
Otherwise, more frequency-dependent terms could enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        Illustration of amplitude birefringence. The GR waveform for the $\ell=|m|=2$ mode of a nonprecessing BBH seen edge-on ($\cos\iota = 0$) is linearly polarized and thus contains equal amounts of left- and right-handed modes for all frequencies (dotted, top).
        However, if spacetime were birefringent (BR) following Eq.~\protect\eqref{eq:waveform_modification}, then the waveform observed on Earth would contain different fractions of the two circular modes, with higher frequencies affected more strongly (solid, top). 
        In the time domain, this manifests as a time-dependent amplification of the waveform, with a stronger effect at later times when the chirp reaches a higher instantaneous frequency (bottom).
        For this example, the black holes do not spin and have equal masses $m_1 = m_2 = 10\, M_\odot$, and we have chosen a luminosity distance $d_L = 400\, {\rm Mpc}$ and $\kappa = 0.6$.
        }
    \label{fig:birefringence}
\end{figure}

@@

To this end, we perform three additional \ac{PE} runs for GW200129, considering only two detectors at a time: LIGO Hanford and Virgo (HV), LIGO Livingston and Virgo (LV), and LIGO Hanford and LIGO Livingston (HL).
If the preference for $\kappa < 0$ is tied to the glitch in Virgo, we expect it to disappear in the HL run, which excludes Virgo data.
